Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-04-06 manuscript/supplemental_material.Rnw
fix linage typo
clean up manuscript and supplemental material

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature,citestyle=numeric-comp,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\usepackage{cleveref}
\usepackage{xspace}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.65\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("magrittr")
library("tidyr")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "placenta_go_results_bp",
                              "placenta_go_results_mf",
                              "hugo_nomenclature",
                              "gene_annotation"
                              ))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
setkey(hugo.nomenclature,"Approved Symbol")
@

@@

  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in}}

@@

    Species                       & Collector & Experiment & Sample(s) \\

@@

    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & SRX1629201 \\
    \textit{Bos taurus}           & Bob Kreft & GSE79121 & SRX1629203 \\
    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & SRX1629204 \\
    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & SRX1629200 \\
    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & SRX1629205 \\

@@

                                               SRR1851012 \\
    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & SRX1629207 \\

@@

                                       SRR243593\\
    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & SRX1629202 \\
    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 \\
    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\

@@

  \caption{Species, experiment and species numbers used in the analyses presented in this paper}

@@

@@

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c"),
             caption="GO terms with significant enrichment ($\\mathrm{FDR}\\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\\% of
  expression. $n(\\mathrm{FDR}\\le0.05)$ is the number of species with
  significant enrichment ($\\mathrm{FDR}\\le 0.05$) in the GO term
  listed. $max(\\mathrm{FDR}\\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.",
  label="tab:go_terms_all_species"
  ),
  math.style.exponents=TRUE,
  include.rownames=FALSE,
  booktabs=TRUE,
  floating=FALSE,
  tabular.environment="longtable",
  caption.placement="top",
  size="\\footnotesize"
  )

@@

@@

<<go.analysis.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 100 \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
    this.label <- paste0("tab:go_terms_enriched_",gsub(" ","_",this.species))
    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=this.label,
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE,
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

@@

<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts.")
    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE, 
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

\section{Gene Expression}

\subsection{Gene Evolution Trees}

\subsubsection{Pre-eclampsia genes}

@@

\subsubsection{Primate Specific Evolution}

@@

@@

\subsection{Clade specific Expression}

@@

\begin{table}[hb]
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{\linagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
\end{table}

\subsection{Species Expression}

<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
    xt.align <- c("r","l","l","p{2.5in}","r")
    if (this.species == "homo sapiens" || this.species == "ateles fusciceps") {
        xt.align <- xt.align[-2]
    }
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 align=xt.align,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@

@@

\begin{table}
\footnotesize
<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
placenta.go.results <- placenta.go.results[,Annotated:=NULL]
setnames(placenta.go.results,"Significant","# Core")
setnames(placenta.go.results,"classic","p value")
setnames(placenta.go.results,"classic.fdr","FDR")
setnames(placenta.go.results,"GO.ID","GO ID")
setnames(placenta.go.results,"Term","GO Term")
print(xtable(placenta.go.results,
             digits=2,
             display=c("d","s","s","d","f","e","e","s")),
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )
@ 
\caption{GO terms most significantly enriched
  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
  in the core placenta transcriptome as compared to all genes
  with 1:1 human orthologs in all species. “\# Core” is the
  number of genes in the core transcriptome for each GO term;
  “Expected” is the number of genes expected in the core
  transcriptome for each term; “p-value” are the odds of this
  occurring by chance calculated by the Fisher exact test for
  each term; “FDR” is the false discovery rate as estimated by
  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
\label{tab:core_placenta_go_terms}
\end{table}

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("min_expression",
             "second_min_expression",
             "third_min_expression",
             "fourth_min_expression",
             "fifth_min_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
           })
cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               gsub("^min_expression$","First",
                                    `Lowest Expression`)]
cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
                                           `Lowest Expression`,perl=TRUE),
                                      levels=c("First","Second","Third","Fourth","Fifth"))
                           ]
setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
      + geom_line(lwd=1)
      + xlab(expression(FPKM))
      + ylab("Genes with N-th Lowest Expression")
      + scale_y_log10()
      + scale_x_log10()
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=100,
                         color=`Lowest Expression`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Lowest Expression`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=100,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Lowest Expression`],
                  show_guide=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Lowest\nExpression"))
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis shows
  the number of orthologous gene sets included in the core placenta
  transcriptome if the least expression cutoff (or second-least, etc.)
  was set to the horizontal axis. For example, if the cutoff is set to 10
  (vertical black line), 37 genes would be included if the FPKM of all
  species must be at least 10, 134 genes would be included if all but
  one species must have FPKM of at least 10. 655 genes would be
  included if all but five species must have FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature,citestyle=numeric-comp,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=blue,citecolor=blue]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\usepackage{cleveref}
\usepackage{xspace}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.65\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("scales")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("magrittr")
library("tidyr")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","hugo_nomenclature","gene_annotation")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "hugo_nomenclature",
                              "gene_annotation"
                              ))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
setkey(hugo.nomenclature,"Approved Symbol")
@

@@

  \small
  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in} c}

@@

    Species                       & Collector & Experiment & Sample(s) & Library \\

@@

    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & SRX1629201 & Total\\
    \textit{Bos taurus}           & Bob Kreft & GSE79121 & SRX1629203 & Total \\
    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & SRX1629204 & Total \\
    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & SRX1629200 & Total\\
    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 & mRNA \\
    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & SRX1629205 & Total \\

@@

                                               SRR1851012  & mRNA \\
    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & SRX1629207 & Total \\

@@

                                       SRR243593 & mRNA \\
    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145 & Total \\
    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & SRX1629202 & Total\\
    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 & Total\\
    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 & Total\\
    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 & mRNA \\

@@

  \caption{Species, experiment and species numbers used in the analyses presented in this paper. 
    The Library column indicates the type of RNA selection performed: Total indicates rRNA depletion; 
    mRNA indicates poly-A selection.}

@@

\section{Species Expression}

<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsubsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
    xt.align <- c("r","l","l","p{2.5in}","r")
    if (this.species == "homo sapiens" || this.species == "ateles fusciceps") {
        xt.align <- xt.align[-2]
    }
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 align=xt.align,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@

@@

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c"),
             caption="GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\\% of
  expression. $n(\\text{FDR}\\le0.05)$ is the number of species with
  significant enrichment ($\\text{FDR}\\le 0.05$) in the GO term
  listed. $max(\\text{FDR}\\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.",
  label="tab:go_terms_all_species"
  ),
  math.style.exponents=TRUE,
  include.rownames=FALSE,
  booktabs=TRUE,
  floating=FALSE,
  tabular.environment="longtable",
  caption.placement="top",
  size="\\footnotesize"
  )

@@

<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts.")
    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE, 
          math.style.exponents=TRUE,
          include.rownames=FALSE,
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

@@

<<go.analysis.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
                           " the top 100 \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
                           " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
    this.label <- paste0("tab:go_terms_enriched_",gsub(" ","_",this.species))
    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=this.label,
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE,
          math.style.exponents=TRUE,
          include.rownames=FALSE,
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

@@

\section{Gene Evolution Trees}

\subsection{Pre-eclampsia genes}

@@

\subsection{Primate Specific Evolution}

@@

\subsection{Other Gene Trees}

@@

\section{Clade specific Expression}

@@

\begin{table}[hb]
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{\lineagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
\end{table}

@@

\begin{table}
\footnotesize
<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis",cache.extra=tools::md5sum(paste0("../data/placenta_go_results_",c("bp","mf","permissive_bp","permissive_mf")))>>=
load("../data/placenta_go_results_mf")
load("../data/placenta_go_results_bp")
load("../data/placenta_go_results_permissive_mf")
load("../data/placenta_go_results_permissive_bp")

placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
placenta.go.results <- placenta.go.results[,Annotated:=NULL]
setnames(placenta.go.results,"Significant","Core")
setnames(placenta.go.results,"classic","p value")
setnames(placenta.go.results,"classic.fdr","FDR")
setnames(placenta.go.results,"GO.ID","GO ID")
setnames(placenta.go.results,"Term","GO Term")

placenta.go.results.permissive <-
    rbindlist(list(placenta_go_results_permissive_mf,
                   placenta_go_results_permissive_bp))
setnames(placenta.go.results.permissive,"Significant","Core")
setnames(placenta.go.results.permissive,"classic","p value")
setnames(placenta.go.results.permissive,"classic.fdr","Permissive FDR")
setnames(placenta.go.results.permissive,"GO.ID","GO ID")
setnames(placenta.go.results.permissive,"Term","GO Term")
setkey(placenta.go.results,"GO ID")
setkey(placenta.go.results.permissive,"GO ID")

combined.placenta.go.results <-
    placenta.go.results.permissive[,list(`GO ID`,`Permissive FDR`)
                                   ][placenta.go.results][order(Ontology,FDR),]
combined.placenta.go.results <-
    combined.placenta.go.results[,`p value`:=as.numeric(`p value`)]
setcolorder(combined.placenta.go.results,
            c("Ontology",
              "GO ID",
              "GO Term",
              "Core",
              "Expected",
              "p value",
              "FDR",
              "Permissive FDR"))
combined.placenta.go.results <-
    combined.placenta.go.results[,`p value`:=NULL]
setnames(combined.placenta.go.results,"Ontology","Ont.")
setnames(combined.placenta.go.results,"Expected","Exp.")
print(xtable(combined.placenta.go.results,
             digits=2,
             display=c("d","s","s","s","d","f","e","e")),
      include.rownames=FALSE,
      booktabs=TRUE,
      math.style.exponents=TRUE,
      floating=FALSE
      )
@ 
\caption{GO terms most significantly enriched
  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP) in
  the core placenta transcriptome ($\text{FPKM}>10$ in 14
  of 14 species) as compared to all genes with 1:1 human orthologs in
  all species. “Ont.” is the GO Ontology, where MF is molecular function 
  and BP is biological process; “Core” is the number of genes in the core
  transcriptome for each GO term; “Exp.” is the number of genes
  expected in the core transcriptome for each term; 
  “FDR” is the false discovery rate as estimated by the
  Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure} based on
  p values calculated by the Fisher Exact Test; 
  “Permissive FDR” is the FDR resulting from the equivalent GO
  analysis using the permissive core placenta transcriptome 
  (1:1 orthologs with expression of at least 10 FPKM at least 10 of 14 species) 
  showing that the enriched GO terms are relatively insensitive to the precise
  set of species included.}
\label{tab:core_placenta_go_terms}
\end{table}

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")
